---
title: "RMS Titanic"
author: "Carter Harris"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|-------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

-   Class: one of `1st`, `2nd`, `3rd`, `Crew`
-   Sex: either `Male` or `Female`
-   Age: either `Child` or `Adult`
-   Survived: either `Yes` or `No`
-   n: integer value, presumably some sort of identification number.

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Write your observations here
    -   The total number of souls on-board the Titanic according to Wikipedia was 2202. The `df-titanic` dataframe contains 2201 entries. As such, the entries probably refer to each individual soul on board the Titanic at the time of sinking.
-   Are there any differences?
    -   Assuming that the dataframe refers to all people on-board the Titanic at the time of sinking, the dataset here includes one less person than is stated to have been on the Titanic, `2201` vs. `2202`.
-   If yes, what might account for those differences?
    -   The Wikipedia article states that the exact number of people that were on-board the Titantic is unknown, given that not everyone who purchased a ticket ended up boarding the vessel. This different in the exact outcomes and exact number of all souls on-board likely accounts for this single-person difference in the total count.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic |>
  filter(Survived == "Yes") |>
  ggplot(aes(x = Class, y = n, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

-   The `Crew` was very heavily skewed `Male` amongst those that survived. I wonder how the `Crew` split more generally; more data is needed on the total number of crew members regardless of their survival.
-   For `1st` class passengers, more than double the number of `Female` passengers survived compared to the `Male` passengers. This split is even more dramatic for the `2nd` class passengers. I again wonder how this compares to the total number of passengers.
-   For the `3rd` class passengers, the split in `Sex` is roughly even, which is very different from that of all other classes.
-   The total number of people who survived seems roughly equal among `1st`, `3rd`, and `Crew` classes while there are overall many less survivors for `2nd` class passengers.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop |>
  filter(Survived == "Yes") |>
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

-   Write your observations here.
    -   For `1st` and `2nd` class passengers, the proportion of passengers who survived appears to be 1, ie all survived. However, we know this isn't actually the case, given that there are also a proportion of passengers in the `No` column.
    -   For `Crew`, the proportion of `Female` passengers that survived was quite high (around 7/8), much higher than that of the `Male` passengers, where less than a quarter survived. This pattern holds true for the `3rd` class passengers, although not to the same extent.
-   Is there anything *fishy* going on in your plot?
    -   Yes! As stated above, all `1st` and `2nd` class passengers seem to have survived, which is know to be untrue.
    -   This is due to the `Adult` and `Child` passenger proportions being computed separetly. When they are graphed and `dodged` together, the proportions for each individually is overlayed. As all the children survived, this proportion of 1 is all that is seen (overlaying over the lower proportion of adults who survived, which is not delineated). To resolve this, the adults and children either need to be shown separately, or the proportions need to be re-computed agnostic to the age of the passengers.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop |>
  filter(Survived == "Yes") |>
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Age)
```

**Observations**:

-   (Write your observations here.)
    -   All of the `1st` and `2nd` class `Child` passengers survived, regardless of `Sex`.
    -   Nearly all the `Female` passengers in `1st` class survived, and a very high percentage of those in `2nd` class did as well. The proportion drops off greatly to less than 50% for the `3rd` class passengers.
    -   Universally, aside from `1st` and `2nd` class children, the `Male` passengers survived at substantially lower rates than that of the `Female` passengers.
    -   `2nd` class `Male`s had the lowest survival rate of any group.
    -   There were no `Child` `Crew`; these are mutually exclusive.
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   As stated above, all `1st/2nd` class `Child` passengers survived; this was overlayed over the lower-proportion survival rates for the `Adult` passengers, as as such, the true data for adult survival rates was not visible in the initial visualization.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
