---
title: "Massachusetts Highway Stops"
author: "Carter Harris"
date: 2025-04-27
output:
  github_document:
    toc: true
---

*Purpose*: In this last challenge we'll focus on using logistic regression to study a large, complicated dataset. Interpreting the results of a model can be challenging---both in terms of the statistics and the real-world reasoning---so we'll get some practice in this challenge.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

*Background*: We'll study data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/), specifically their dataset on Massachusetts State Patrol police stops.

```{r setup}
library(tidyverse)
library(broom)
```

# Setup

<!-- -------------------------------------------------- -->

### **q1** Go to the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/) page and download the Massachusetts State Police records in `Rds` format. Move the data to your `data` folder and match the `filename` to load the data.

*Note*: An `Rds` file is an R-specific file format. The function `readRDS` will read these files.

```{r q1-task}
## TODO: Download the data, move to your data folder, and load it
filename <- "data/snf_opp_ma.rds"
df_data <- readRDS(filename)
```

# EDA

<!-- -------------------------------------------------- -->

### **q2** Do your "first checks" on the dataset. What are the basic facts about this dataset?

```{r q2-task}
head(df_data)
summary(df_data)
glimpse(df_data)
```

**Observations**:

-   What are the basic facts about this dataset?
    -   This dataset is quite old (for this space) with the earliest data from 2007 and newest data from 2015.
    -   Given this is police data and the type for all is "vehicular," I assume this is traffic stop data.
    -   The data contains lots of information about the person who was stopped (race, age, sex).
    -   The data contains outcomes about the stop (arrest_made, citation issued, warning issued)
    -   The data contians information about what happened during the stop (search conducted, contraband found, frisk performed, etc).
    -   There does seem to be a lot of holes and NAs in this dataset.

Note that we have both a `subject_race` and `race_Raw` column. There are a few possibilities as to what `race_Raw` represents:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

Let's try to distinguish between these two possibilities.

### **q3** Check the set of factor levels for `subject_race` and `raw_Race`. What do you note about overlap / difference between the two sets?

```{r q3-task}
## TODO: Determine the factor levels for subject_race and raw_Race
df_data %>% 
  pull(subject_race) %>% 
  levels()

print("\n")

df_data %>% 
 pull(raw_Race) %>% 
  unique()
```

Note - I left one commented out to make it easier to distinguish between the two.

**Observations**:

-   What are the unique values for `subject_race`?
    -   `subject_race` contains:
        -   `asian/pacific islander`
        -   `black`
        -   `hispanic`
        -   `white`
        -   `other`
        -   and `unknown`
-   What are the unique values for `raw_Race`?
    -   `raw_Race` contains:
        -   `White`
        -   `Black`
        -   `Middle Eastern or East Indian (South Asian)`
        -   `Hispanic`
        -   `Asian or Pacific Islander`
        -   `American Indian or Alaskan Native`
        -   `None - for no operator present citations only`
        -   `NA`
-   What is the overlap between the two sets?
    -   White, black, Hispanic, Asian/PI, and a catchall category are included for both.
-   What is the difference between the two sets?
    -   `raw_Race` contains several additional categories that are not included in `subject_race`, such as Middle Eastern/East Indian and American Indian. Presumably, these were lumped into additional categories or other to reduce levels in `subject_race`. Additionally, the `None - for no operator present citations only` were presumably lumped in with unknown or excluded from `subject_race`.

### **q4** Check whether `subject_race` and `raw_Race` match for a large fraction of cases. Which of the two hypotheses above is most likely, based on your results?

*Note*: Just to be clear, I'm *not* asking you to do a *statistical* hypothesis test.

```{r q4-task}
## TODO: Devise your own way to test the hypothesis posed above.
df_data %>% 
  mutate(race_match = (subject_race == tolower(raw_Race))) %>% 
  summarise(percent = mean(race_match, na.rm = TRUE) * 100)
```

**Observations**

Between the two hypotheses:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

which is most plausible, based on your results?

-   `race_Raw` is an unprocessed version of `subject_race`

## Vis

<!-- ------------------------- -->

### **q5** Compare the *arrest rate*---the fraction of total cases in which the subject was arrested---across different factors. Create as many visuals (or tables) as you need, but make sure to check the trends across all of the `subject` variables. Answer the questions under *observations* below.

(Note: Create as many chunks and visuals as you need)

```{r vis1}
df_data %>% 
  group_by(subject_age) %>% 
  summarise(arrest_mean = mean(arrest_made, na.rm = TRUE)) %>% 
  ggplot(aes(subject_age, arrest_mean)) +
  geom_point() +
  geom_smooth() +
  labs(
    x = "Age (years)",
    y = "Percent of Stops Resulting in Arrests",
    title = "Age vs. Arrest Rate"
  )
```

```{r vis2}
df_data %>%
  drop_na(subject_sex) %>% 
  group_by(subject_sex) %>% 
  summarise(arrest_mean = mean(arrest_made, na.rm = TRUE)) %>% 
  ggplot(aes(subject_sex, arrest_mean, fill=subject_sex)) +
  geom_col() +
  labs(
    x = "Sex",
    y = "Percent of Stops Resulting in Arrests",
    title = "Sex vs. Arrest Rate"
  )
```

```{r}
df_data %>% 
  drop_na(subject_race) %>% 
  group_by(subject_race) %>% 
  summarise(arrest_mean = mean(arrest_made, na.rm = TRUE)) %>% 
  ggplot(aes(subject_race, arrest_mean, fill=subject_race)) +
  geom_col() +
  labs(
    x = "Race",
    y = "Percent of Stops Resulting in Arrests",
    title = "Race vs. Arrest Rate"
  )
```

**Observations**:

-   How does `arrest_rate` tend to vary with `subject_age`?
    -   There is a rather clear trend in arrest rate by age. Arrest rate peaks in the late 20s-early 30s and steadily decreases as age increases.
-   How does `arrest_rate` tend to vary with `subject_sex`?
    -   Men are arrested at a rate twice that of women.
-   How does `arrest_rate` tend to vary with `subject_race`?
    -   Most races are fairly similar, although hispanic people are more than twice as likely as almost every other group to be arrested.

# Modeling

<!-- -------------------------------------------------- -->

We're going to use a model to study the relationship between `subject` factors and arrest rate, but first we need to understand a bit more about *dummy variables*

### **q6** Run the following code and interpret the regression coefficients. Answer the the questions under *observations* below.

```{r q6-task}
## NOTE: No need to edit; inspect the estimated model terms.
fit_q6 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q6 %>% tidy()
```

**Observations**:

-   Which `subject_race` levels are included in fitting the model?
    -   `subject_race` only includes arrests where the race is white, black, or hispanic.
-   Which `subject_race` levels have terms in the model?
    -   It appears that only the hispanic and white level is included in the model (ie, black is missing).

You should find that each factor in the model has a level *missing* in its set of terms. This is because R represents factors against a *reference level*: The model treats one factor level as "default", and each factor model term represents a change from that "default" behavior. For instance, the model above treats `subject_sex==male` as the reference level, so the `subject_sexfemale` term represents the *change in probability* of arrest due to a person being female (rather than male).

The this reference level approach to coding factors is necessary for [technical reasons](https://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture10/lecture10-94842.html#why-is-one-of-the-levels-missing-in-the-regression), but it complicates interpreting the model results. For instance; if we want to compare two levels, neither of which are the reference level, we have to consider the difference in their model coefficients. But if we want to compare all levels against one "baseline" level, then we can relevel the data to facilitate this comparison.

By default `glm` uses the first factor level present as the reference level. Therefore we can use `mutate(factor = fct_relevel(factor, "desired_level"))` to set our `"desired_level"` as the reference factor.

### **q7** Re-fit the logistic regression from q6 setting `"white"` as the reference level for `subject_race`. Interpret the the model terms and answer the questions below.

```{r q7-task}
## TODO: Re-fit the logistic regression, but set "white" as the reference
## level for subject_race
df_data_reorder <-
  df_data %>% 
  mutate(subject_race = fct_relevel(subject_race, "white"))

fit_q7 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data_reorder %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q7 %>% tidy()
```

**Observations**:

-   Which `subject_race` level has the highest probability of being arrested, according to this model? Which has the lowest probability?
    -   Hispanic has the highest probability according to this model based upon having the highest estimated value. Being white has the lowest probability (of the three tested) given it results in a negative estimate coefficient.
-   What could explain this difference in probabilities of arrest across race? List **multiple** possibilities.
    -   It is hard to speculate why these disparities exist without knowing the reason for arrest. Are certain races stopped for different reasons?
-   Look at the set of variables in the dataset; do any of the columns relate to a potential explanation you listed?
    -   The presence of contraband and whether different races were more likely to have contraband could help explain this behavior.

One way we can explain differential arrest rates is to include some measure indicating the presence of an arrestable offense. We'll do this in a particular way in the next task.

### **q8** Re-fit the model using a factor indicating the presence of contraband in the subject's vehicle. Answer the questions under *observations* below.

```{r q8-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop

fit_q8 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )


fit_q8 %>% tidy()
```

**Observations**:

-   How does controlling for found contraband affect the `subject_race` terms in the model?
    -   The influence of race decreases considerably and is dwarfed by the intensity of found contraband in terms of determining arrest.
-   What does the *finding of contraband* tell us about the stop? What does it *not* tell us about the stop?
    -   Although the data on what type on contraband exists, it is not accounted for in this simpler model. Further, we don't know how race or other factors influence the probability of a search being conducted in the first place.

### **q9** Go deeper: Pose at least one more question about the data and fit at least one more model in support of answering that question.

**Observations**:

-   Document your question and findings

## Further Reading

<!-- -------------------------------------------------- -->

-   Stanford Open Policing Project [findings](https://openpolicing.stanford.edu/findings/).
